Extended Data

### 2022-02-21 doc/manuscript/Dow_manuscript_Nature.Rmd
Update Dow_manuscript_Nature.Rmd
start replacing references to growth "phenology"

Tree-ring records, which can be used to examine relationships of annual growth to temperature but not to understand growth phenology, reveal that growth of temperate deciduous trees tends to be most sensitive to temperature or potential evapotranspiration between late spring and early summer[@dorangeville_drought_2018;@helcoski_growing_2019]<!-- dorangeville_drought_2018 looks at PET.-->, with some hints that warmer springs may have a modest positive effect on growth[@helcoski_growing_2019].<!-- add more refs here, but only if already cited elsewhere (to keep n references down-->
Thus, tree-ring evidence does not necessarily align with the finding that warming spring temperatures increase annual forest CO~2~ uptake[@keenan_net_2014].
Characterizing responses of stem growth to warming spring temperatures is critical to bridging this conceptual disconnect and understanding how forest biomass growth is likely to change as the climate warms.

@@

We fit a five-parameter logistic growth model[@mcmahon_general_2015] to dendrometer band data from each tree-year to define stem growth milestones and growth rates (Fig. `r schematic_summary`). In particular, we model the observed diameter at breast height (DBH) on a given day of the year (DOY; *i.e.*, julian days) as: 
$$ DBH = L + \frac{K - L} { 1 + 1/\theta \cdot exp[-r(DOY - DOY_{ip}) / \theta)] ^ \theta}$$ 
Here, $L$ and $K$ are lower and upper asymptotes of the model. 
$DOY_{ip}$ is the day of year where the inflection point in growth rate occurs,
$r$ shapes the slope of the curve at the inflection point,
and $\theta$ is a tuning parameter controlling the slope of the curve toward the upper asymptote.
This allows an asymetric fit to the data, where the onset of growth can be estimated independent of the cessation of growth.
The DOY on which maximum growth occurs, $DOY_{g_{max}}$ (Fig. `r schematic_summary`), occurs on $DOY_{ip}$ when $\theta = 1$.
The model outputs two additional variables, $a$ and $b$, which represent the beginning and end DBH in each model year. We used the first and last manual dendrometer band measurements as starting points to calculate these two variables. With this method, initial measurements too late in the growing season which miss the start of the growing season can cause incorrect estimations of $a$ and $b$. The 1998 survey at Harvard Forest recorded first measurements on May 26, which is less than 2 weeks away from the average DOY25 at this site. This late start made the calculation of timing and total growth variables unreliable, so we removed tree years in 1998 from our analysis. 
The model was fit in R v4.0 using the functions developed in the *Rdendrom* package[@mcmahon_general_2015]. These functions take the time-series of manual dendrometer band measurements and return maximum-likelihood optimized values of the above five parameters that best predict DBH for each day of year. 
An advantage of this approach is that short-term shrinkage and swelling associated with rain events [@mcmahon_general_2015; @herrmann_tree_2016] and measurement errors show up as residual variation and do not unduly influence the parameters of interest. 
Having fit the model for each band-year of data, we then modeled DBH using these optimal parameter values in our logistic growth model and extracted the intra-annual growth variables of interest (Fig. `r schematic_summary`).

@@

<!-- Determining critical temperature windows -->
The critical temperature window (CTW, Fig. `r schematic_summary`), defined as the period over which $T_{max}$ was most strongly correlated with $DOY_{25}$, was determined using the R package *climwin*[@vandepol_identifying_2016]. 
This package tests the correlation between one or more predictor climate variable and a biological outcome variable over all consecutive time windows within a specified time-frame. It does so by reporting the correlation and $\Delta AICc$, the difference in Akaike Information Criterion corrected for small sample size relative to a null model for each window. 
Here, we tested for correlation between temperature predictor variables ($T_{max}$, $T_{min}$) and biological outcome variable $DOY_{25}$ over the time-frame from January 1 to the mean $DOY_{25}$ for the species group (by xylem porosity) and site (`r  EDT_phenology_parameters`). 
The time period yielding the lowest $\Delta AICc$ was selected as the CTW.
Because $T_{max}$ proved to have a generally stronger influence over $DOY_{25}$ and other growth parameters, we focused on this variable in our ultimate model, as opposed to $T_{min}$.
We defined CTW for $DOY_{25}$, as opposed to other growth phenology parameters, because spring temperatures should have the most direct influence on this variable.

>>>>

Tree-ring records, which can be used to examine relationships of annual growth to temperature but not to understand growth seasonality, reveal that growth of temperate deciduous trees tends to be most sensitive to temperature or potential evapotranspiration between late spring and early summer[@dorangeville_drought_2018;@helcoski_growing_2019]<!-- dorangeville_drought_2018 looks at PET.-->, with some hints that warmer springs may have a modest positive effect on growth[@helcoski_growing_2019].<!-- add more refs here, but only if already cited elsewhere (to keep n references down-->
Thus, tree-ring evidence does not necessarily align with the finding that warming spring temperatures increase annual forest CO~2~ uptake[@keenan_net_2014].
Characterizing responses of stem growth to warming spring temperatures is critical to bridging this conceptual disconnect and understanding how forest biomass growth is likely to change as the climate warms.

@@

We fit a five-parameter logistic growth model[@mcmahon_general_2015] to dendrometer band data from each tree-year to define stem growth milestones and growth rates (Fig. `r schematic_summary`). In particular, we model the observed diameter at breast height (DBH) on a given day of the year (DOY; *i.e.*, julian days) as: 
$$ DBH = L + \frac{K - L} { 1 + 1/\theta \cdot exp[-r(DOY - DOY_{ip}) / \theta)] ^ \theta}$$ 
Here, $L$ and $K$ are lower and upper asymptotes of the model. 
$DOY_{ip}$ is the day of year where the inflection point in growth rate occurs,
$r$ shapes the slope of the curve at the inflection point,
and $\theta$ is a tuning parameter controlling the slope of the curve toward the upper asymptote.
This allows an asymmetric fit to the data, where the onset of growth can be estimated independent of the cessation of growth.
The DOY on which maximum growth occurs, $DOY_{g_{max}}$ (Fig. `r schematic_summary`), occurs on $DOY_{ip}$ when $\theta = 1$.
The model outputs two additional variables, $a$ and $b$, which represent the beginning and end DBH in each model year. We used the first and last manual dendrometer band measurements as starting points to calculate these two variables. With this method, initial measurements too late in the growing season which miss the start of the growing season can cause incorrect estimations of $a$ and $b$. The 1998 survey at Harvard Forest recorded first measurements on May 26, which is less than 2 weeks away from the average DOY25 at this site. This late start made the calculation of timing and total growth variables unreliable, so we removed tree years in 1998 from our analysis. 
The model was fit in R v4.0 using the functions developed in the *Rdendrom* package[@mcmahon_general_2015]. These functions take the time-series of manual dendrometer band measurements and return maximum-likelihood optimized values of the above five parameters that best predict DBH for each day of year. 
An advantage of this approach is that short-term shrinkage and swelling associated with rain events [@mcmahon_general_2015; @herrmann_tree_2016] and measurement errors show up as residual variation and do not unduly influence the parameters of interest. 
Having fit the model for each band-year of data, we then modeled DBH using these optimal parameter values in our logistic growth model and extracted the intra-annual growth variables of interest (Fig. `r schematic_summary`).

@@

<!-- Determining critical temperature windows -->
The critical temperature window (CTW, Fig. `r schematic_summary`), defined as the period over which $T_{max}$ was most strongly correlated with $DOY_{25}$, was determined using the R package *climwin*[@vandepol_identifying_2016]. 
This package tests the correlation between one or more predictor climate variable and a biological outcome variable over all consecutive time windows within a specified time-frame. It does so by reporting the correlation and $\Delta AICc$, the difference in Akaike Information Criterion corrected for small sample size relative to a null model for each window. 
Here, we tested for correlation between temperature predictor variables ($T_{max}$, $T_{min}$) and biological outcome variable $DOY_{25}$ over the time-frame from January 1 to the mean $DOY_{25}$ for the species group (by xylem porosity) and site (`r  EDT_phenology_parameters`). 
The time period yielding the lowest $\Delta AICc$ was selected as the CTW.
Because $T_{max}$ proved to have a generally stronger influence over $DOY_{25}$ and other growth parameters, we focused on this variable in our ultimate model, as opposed to $T_{min}$.
We defined CTW for $DOY_{25}$, as opposed to other parameters describing the timing of growth, because spring temperatures should have the most direct influence on this variable.
